Bumble bee foraging bias

### 2021-01-13 code/visitation_bias.Rmd
restructured project directory; this will break some paths, but easy to fix, and clearer now

---
title: "visitation_bias"
author: "Doug Sponsler"
date: "12/14/2020"
output: html_document
---
# Preparations
## Packages 
```{r, warning=FALSE, include=FALSE, echo = FALSE}
library(bipartite)
library(mgcv)
library(mgcViz)
library(brms)
library(lme4)
#library(inla)
library(bayesplot)
library(vegan)
#library(tidybayes)
library(tidyverse)
library(lubridate)
library(tidymodels)
```

@@

# 1. Is bumble bee foraging biased? How do taxonomic, phylogenetic, and morphological bias compare -- NO, can't analyze this, these should be considered separate repsonse variables and analyzed in separate models? Are some BB species more biased than others?
## 1.1 Calculate Blüthgen's d' (i.e. visitation bias)
```{r}
# No grouping -- just plant.sp
bias_plantsp <- biaseval(visit_cover, plant.sp, site, date) %>%
  left_join(site_data) %>% # add site data
  left_join(bb_abundance) %>% # add BB abundance data
  left_join(fl_abundance) %>% # add fl abundance
  mutate(criterion = factor(rep("plant.sp", n())))

@@

ggplot(bias_plantsp.sitedate, aes(log(bb.sp.prop.abund), dprime)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~bb.sp)
```

## 1.3 Multi-level model: bias ~ elevation + (1 + elevation | bb.sp)
```{r}
# brms
brm.ktype1 <- brm(dprime ~ yday + elev + (1 + elev | bb.sp), 
                  control = list(max_treedepth = 10),
                  cores = 3,
                  iter = 2000,
                  file = "./output/brm_ktype1",
                  data = filter(bias, criterion ==  "ktype"))

summary(brm.ktype1)
plot(brm.ktype1)
parnames(brm.ktype1)
mcmc_plot(brm.ktype1, pars = c("^r_"))
mcmc_plot(brm.ktype1, pars = c("^b_"))
mcmc_plot(brm.ktype1)
conditional_effects(brm.ktype1)
pp_check(brm.ktype1)

brm.ktype2 <- brm(dprime ~ 1 + yday + elev + (1 | bb.sp), 
                  control = list(max_treedepth = 10),
                  cores = 4,
                  iter = 2000,
                  file = "./output/brm_ktype2b",
                  data = filter(bias, criterion ==  "ktype"))

summary(brm.ktype2)
plot(brm.ktype2)
parnames(brm.ktype2)
mcmc_plot(brm.ktype2, pars = c("^r_"))
mcmc_plot(brm.ktype2, pars = c("^b_"))
mcmc_plot(brm.ktype2)
conditional_effects(brm.ktype2)
pp_check(brm.ktype2)

brm.ktype3 <- brm(dprime ~ 1 + (1 | bb.sp), 
                  control = list(max_treedepth = 10),
                  cores = 4,
                  iter = 2000,
                  family = "poisson",
                  file = "./output/brm_ktype3",
                  data = filter(bias, criterion ==  "ktype"))

summary(brm.ktype3)
plot(brm.ktype3)
parnames(brm.ktype3)
mcmc_plot(brm.ktype3, pars = c("^r_"))
mcmc_plot(brm.ktype3, pars = c("^b_"))
mcmc_plot(brm.ktype3)
conditional_effects(brm.ktype3)
pp_check(brm.ktype3)

# lme4
library(lme4)
library(ggeffects)

lmer.ktype <- lmer(dprime ~ 1 + elev:yday + (1 | bb.sp), 
                  data = filter(bias, criterion ==  "ktype"))
summary(lmer.ktype)
pred.ktype_yday <- ggpredict(lmer.ktype, "yday")
pred.ktype_elev <- ggpredict(lmer.ktype, "elev")
#pred.ktype_yday_ran <- ggpredict(lmer.ktype, "yday", type = "random")
#pred.ktype_elev_ran <- ggpredict(lmer.ktype, "elev", type = "random")
pred.ktype_bbsp <- ggpredict(lmer.ktype, "bb.sp", type = "random")

plot(pred.ktype_yday)
plot(pred.ktype_elev)
#plot(pred.ktype_yday_ran)
#plot(pred.ktype_elev_ran)
plot(pred.ktype_bbsp)

# mgcv
gam.ktype <- bam(dprime ~ s(bb.sp, bs = "re") + 
                   t2(yday, elev.mean, bs = c("tp", "gp")),
                 method = "fREML",
                 discrete = TRUE,
                 data = filter(bias, criterion ==  "ktype")) %>% getViz()

summary(gam.ktype)
plot(gam.ktype, allTerms = TRUE)

# inla
library(INLA)

inla.ktype <- inla(dprime ~ elev.mean, 
                   family = "gaussian",
                   data = filter(bias, criterion ==  "ktype"))

inla.pen <- inla(dprime ~ 1 + elev.mean,
  data = filter(bias, criterion ==  "ktype"), control.predictor = list(compute = TRUE))
summary(inla.pen)
```

@@

1. Bias could  be trated simply as the Euclidean difference between prop.flower.cover and prop.visits. This has the effect of weighting bias by flower.cover, particularly negative bias, since an abundant plant can have a greater negative bias (because it is farther from 0) than a rare plant. Euclidean bias is bounded between -1 and 1. Its symmetry is a nice property that sets it apart from proportional or fold-change bias. If, for whatever reason, I decided that a log-transformation of this measure were desirable, I think I could simply add a constant of 1 to all values, which would move the boundary up to (0, 2). I think the biggest problem with measuring bias in this way is that it makes no distinction between a plant that was visited a little and one that was visited not at all, and that could be an important biological difference. For example, consider two plants, one with a proportional abundance = 0.01 and proportional visitation = 0.01, and the second with proportional abundance = 0.01 and proportional visitation = 0.00. These will look identical in our analysis, but it is possible that the unvisited plant would not have been visited no matter how high its abundance was. **For starters, though, I am going to proceed with this metric because it is the simplest. And because DJ thought it was a good idea.**

@@

## 2.3 Calculate bias per plant speces/ktype/family
```{r}
bias_group_plantsp <- biaseval_group(visit_cover, plant.sp) %>%
  left_join(site_data) %>%
  rename(group = criterion) %>%
  mutate(criterion = rep("plant.sp", n()))

bias_group_ktype <- biaseval_group(visit_cover, k.type.ss) %>%
  left_join(site_data) %>%
  rename(group = criterion) %>%
  mutate(criterion = rep("ktype", n()))

bias_group_family <- biaseval_group(visit_cover, plant.family) %>%
  left_join(site_data) %>%
  rename(group = criterion) %>%
  mutate(criterion = rep("plant.family", n()))

bias_group <- bind_rows(bias_group_plantsp, bias_group_ktype, bias_group_family)

ggplot(bias_group, aes(prop.flower.cover, prop.visits)) +
  geom_point(alpha = 0.1) +
  #geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
  geom_smooth(method = "lm") +
  geom_abline(slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~criterion)

ggplot(bias_group, aes(log(prop.flower.cover), prop.visits)) +
  geom_point(alpha = 0.1) +
  geom_smooth() +
  facet_wrap(~criterion)

ggplot(filter(bias_group, criterion == "ktype"), aes(prop.flower.cover, prop.visits)) +
  geom_point(alpha = 0.1) +
  geom_smooth() +
  geom_abline(slope = 1, color = "red", linetype = "dashed") +
  facet_wrap(~group)
```

@@

## 2.5 Explaining bias by floral traits, BB spp, and elevation (and maybe time)
```{r}
repmod.ktype <- brm(bias ~ elev.mean + (1 | criterion), 
                    data = bias_ktype)

repmod.ktype <- brm(bias ~ criterion, 
                    data = bias_ktype)
parnames(repmod.ktype)
mcmc_plot(repmod.ktype)
conditional_effects(repmod.ktype)

H.K5_K6 <- paste("b_criterion5 = b_criterion6")
hypothesis(repmod.ktype, H.K5_K6, class = "")
```

# How do species vary in their biases
```{r}
bias_matrix.plantsp <- bias_group_plantsp %>%
  group_by(bb.sp, criterion) %>%
  summarize(bias = mean(bias)) %>%
  pivot_wider(values_from = bias, names_from = criterion, values_fill = 0) %>%
  column_to_rownames("bb.sp")

bias_matrix.ktype <- bias_group_ktype %>%
  group_by(bb.sp, criterion) %>%
  summarize(bias = mean(bias)) %>%
  pivot_wider(values_from = bias, names_from = criterion, values_fill = 0) %>%
  column_to_rownames("bb.sp")

library(vegan)

pca.plantsp <- rda(bias_matrix.plantsp)
plot(pca.plantsp, type = "t")

pca.ktype <- rda(bias_matrix.ktype)
plot(pca.ktype)

```

>>>>

---
title: "visitation_bias"
author: "Doug Sponsler"
date: "12/14/2020"
output: html_document
---
# Preparations
## Packages 
```{r, warning=FALSE, include=FALSE, echo = FALSE}
library(bipartite)
library(mgcv)
library(mgcViz)
library(brms)
library(lme4)
library(ggeffects)
library(sjPlot)
#library(inla)
library(bayesplot)
library(vegan)
#library(tidybayes)
library(tidyverse)
library(lubridate)
library(tidymodels)
```

@@

# 1. How does overall bias (Blüthgen's d') vary across species and through elev*time? For now, we will focus on bias at the level of plant species. We'll start considering traits later on when our goal is to explain patterns of bias.
## 1.1 Calculate Blüthgen's d' (i.e. visitation bias)
```{r}
# No grouping -- just plant.sp
bias_plantsp <- biaseval(visit_cover, plant.sp, site, date) %>%
  left_join(site_data) %>% # add site data
  left_join(bb_abundance) %>% # add BB abundance data
  left_join(fl_abundance) %>% # add fl abundance
  mutate(criterion = factor(rep("plant.sp", n())))

@@

ggplot(bias, aes(log(bb.sp.prop.abund), dprime)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~bb.sp)
```

## 1.3 Multi-level model: bias ~ yday * elevation + (1 + elevation | bb.sp)
```{r}
# I am interested both in the variation of bias through elevation and time and in the intercept effect of each bumble bee species

# First, let's look at the distribution of bias per BB sp
ggplot(filter(bias, criterion == "plant.sp"), aes(dprime)) +
  geom_histogram() +
  facet_wrap(~bb.sp)

# Now, let's try to get a sense for the shape of the response to elevation and time; we might expect to find nonlinearities here
ggplot(filter(bias, criterion == "plant.sp"), aes(elev, dprime)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5)) +
  facet_wrap(~bb.sp)

ggplot(filter(bias, criterion == "plant.sp"), aes(yday, dprime)) +
  geom_point() +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5)) +
  facet_wrap(~bb.sp)

# After inspecting these plots, I see no reason to indulge in wiggly models.

# This model includes a global interaction term for yday and elevation, with a varying intercept for bb.sp and a varying slow for elev | bb.sp. I think this should give me all the estimation I'm interested in.
brm.plantsp11 <- brm(dprime ~ yday * elev + (1 + elev | bb.sp), 
                  control = list(max_treedepth = 10),
                  cores = 4,
                  iter = 4000,
                  #family = "skew_normal",
                  file = "./output/brm_plantsp11",
                  data = filter(bias, criterion ==  "plant.sp"))

summary(brm.plantsp11)
prior_summary(brm.plantsp11)
plot(brm.plantsp11)
parnames(brm.plantsp11)
mcmc_plot(brm.plantsp11, pars = c("^r_"))
mcmc_plot(brm.plantsp11, pars = c("elev"))
mcmc_plot(brm.plantsp11, pars = c("^b_"))
mcmc_plot(brm.plantsp11)
conditional_effects(brm.plantsp11)
pp_check(brm.plantsp11)
bayes_R2(brm.plantsp11)
hypothesis(brm.plantsp11, "b_elev = 0", class = NULL)

brm.plantsp1 <- brm(dprime ~ yday * elev + (1 | bb.sp), 
                  control = list(max_treedepth = 10),
                  cores = 4,
                  iter = 2000,
                  file = "./output/brm_plantsp1",
                  data = filter(bias, criterion ==  "plant.sp"))

summary(brm.plantsp1)
prior_summary(brm.plantsp1)
plot(brm.plantsp1)
parnames(brm.plantsp1)
mcmc_plot(brm.plantsp1, pars = c("^r_"))
mcmc_plot(brm.plantsp1, pars = c("elev"))
mcmc_plot(brm.plantsp1, pars = c("^b_"))
mcmc_plot(brm.plantsp1)
conditional_effects(brm.plantsp1)
pp_check(brm.plantsp1)
bayes_R2(brm.plantsp1)
hypothesis(brm.plantsp1, "b_elev = 0", class = NULL)

brm.plantsp2 <- brm(dprime ~ yday + elev + (1 | bb.sp), 
                  control = list(max_treedepth = 10),
                  cores = 4,
                  iter = 2000,
                  file = "./output/brm_plantsp2",
                  data = filter(bias, criterion ==  "plant.sp"))

summary(brm.plantsp2)
prior_summary(brm.plantsp2)
plot(brm.plantsp2)
parnames(brm.plantsp2)
mcmc_plot(brm.plantsp2, pars = c("^r_"))
mcmc_plot(brm.plantsp2, pars = c("elev"))
mcmc_plot(brm.plantsp2, pars = c("^b_"))
mcmc_plot(brm.plantsp2)
conditional_effects(brm.plantsp2)
pp_check(brm.plantsp2)
bayes_R2(brm.plantsp2)
hypothesis(brm.plantsp2, "b_elev = 0", class = NULL)

# This, I think, would also be a legitimate way to specify the model, but it would not tell us anything about elevation or time. 
brm.plantsp0 <- brm(dprime ~ 1 + (1 | bb.sp) + (1 | site),
                  control = list(max_treedepth = 10),
                  cores = 4,
                  iter = 2000,
                  file = "./output/brm_plantsp0",
                  data = filter(bias, criterion ==  "plant.sp"))

summary(brm.plantsp0)
plot(brm.plantsp0)
parnames(brm.plantsp0)
mcmc_plot(brm.plantsp0, pars = c("^r_"))
mcmc_plot(brm.plantsp0, pars = c("^b_"))
pp_check(brm.plantsp0)
bayes_R2(brm.plantsp0)
```

@@

1. Bias could  be treated simply as the Euclidean difference between prop.flower.cover and prop.visits. This has the effect of weighting bias by flower.cover, particularly negative bias, since an abundant plant can have a greater negative bias (because it is farther from 0) than a rare plant. Euclidean bias is bounded between -1 and 1. Its symmetry is a nice property that sets it apart from proportional or fold-change bias. If, for whatever reason, I decided that a log-transformation of this measure were desirable, I think I could simply add a constant of 1 to all values, which would move the boundary up to (0, 2). I think the biggest problem with measuring bias in this way is that it makes no distinction between a plant that was visited a little and one that was visited not at all, and that could be an important biological difference. For example, consider two plants, one with a proportional abundance = 0.01 and proportional visitation = 0.01, and the second with proportional abundance = 0.01 and proportional visitation = 0.00. These will look identical in our analysis, but it is possible that the unvisited plant would not have been visited no matter how high its abundance was. **For starters, though, I am going to proceed with this metric because it is the simplest. And because DJ thought it was a good idea.**

@@

## 2.3 Calculate bias per plant speces/ktype/family
```{r}
representation <- biaseval_group(visit_cover, plant.sp) %>%
  rename(plant.sp = criterion) %>%
  left_join(site_data) %>%
  left_join(fl_traits) %>%
  mutate(k.type.ss = factor(k.type.ss)) %>%
  select(site, date, elev.mean, yday, bb.sp, plant.sp, k.type.ss, 
         prop.visits, prop.flower.cover, bias, visits, 
         flower.cover, plant.genus, plant.family)

# 
representation_ktype <- biaseval_group(visit_cover, k.type.ss) %>%
  left_join(site_data) %>%
  rename(group = criterion) %>%
  mutate(criterion = rep("ktype", n()))
# 
# bias_group_family <- biaseval_group(visit_cover, plant.family) %>%
#   left_join(site_data) %>%
#   rename(group = criterion) %>%
#   mutate(criterion = rep("plant.family", n()))
# 
# bias_group <- bind_rows(bias_group_plantsp, bias_group_ktype, bias_group_family)
# 
# ggplot(bias_group, aes(prop.flower.cover, prop.visits)) +
#   geom_point(alpha = 0.1) +
#   #geom_smooth(method = "gam", formula = y ~ s(x, k = 4)) +
#   geom_smooth(method = "lm") +
#   geom_abline(slope = 1, color = "red", linetype = "dashed") +
#   facet_wrap(~criterion)
# 
# ggplot(bias_group, aes(log(prop.flower.cover), prop.visits)) +
#   geom_point(alpha = 0.1) +
#   geom_smooth() +
#   facet_wrap(~criterion)
# 
# ggplot(filter(bias_group, criterion == "ktype"), aes(prop.flower.cover, prop.visits)) +
#   geom_point(alpha = 0.1) +
#   geom_smooth() +
#   geom_abline(slope = 1, color = "red", linetype = "dashed") +
#   facet_wrap(~group)
```

@@

## 2.5 Explaining bias by floral traits, BB spp, and elevation (and maybe time)
```{r}
ggplot(representation, aes(k.type.ss, bias)) +
  geom_violin()

ggplot(representation_ktype, aes(group, bias)) +
  geom_boxplot()

ggplot(representation, aes(bias)) +
  geom_histogram() +
  facet_wrap(~k.type.ss)

ggplot(filter(representation, visits > 0), aes(bias)) +
  geom_histogram() +
  facet_wrap(~k.type.ss)

ggplot(filter(representation, prop.flower.cover > 0.05), aes(k.type.ss, bias)) +
  geom_boxplot()

ggplot(representation, aes(prop.flower.cover, prop.visits)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~k.type.ss)

summary(lm(representation ~ k.type.ss, data = bias))

brm_representation0 <- brm(bias ~ 1 + (1 | group),
                 control = list(max_treedepth = 10),
                 cores = 4,
                 iter = 2000,
                 file = "./output/brm_representation1",
                 data = representation_ktype)

brm_representation1 <- brm(bias ~ 1 + (1 | k.type.ss),
                 control = list(max_treedepth = 10),
                 #cores = 4,
                 cores = 4,
                 chains = 4,
                 #iter = 2000,
                 iter = 2000,
                 file = "./output/brm_representation1",
                 data = representation)

summary(brm_representation1)
plot(brm_representation1)
parnames(brm_representation1)
mcmc_plot(brm_representation1, pars = c("^r_"))
#mcmc_plot(representation.ktype1, pars = c("elev"))
mcmc_plot(brm_representation1, pars = c("^b_"))
mcmc_plot(brm_representation1)
#conditional_effects(representation.ktype1)
pp_check(brm_representation1)
bayes_R2(brm_representation1)

hypothesis(representation.ktype1, "r_group[5,Intercept] = r_group[6,Intercept]", class = "")
hypothesis(representation.ktype1, "r_group[1,Intercept] = r_group[0,Intercept]", class = "")

# One interesting way to think about bias is to focus on the slope of prop.visits ~ prop.flower.cover. 
# We could analyze this with a varying slope model

# Binomial ggplot smoother
binomial_smooth <- function(...) {
  geom_smooth(method = "glm", method.args = list(family = "quasibinomial"), se = FALSE, ...)
}

ggplot(representation, aes(sqrt(prop.flower.cover), sqrt(prop.visits))) +
  geom_point() +
  #geom_smooth(method = "lm") +
  #binomial_smooth(formula = y ~ splines::ns(x, 3)) +
  geom_smooth() +
  #geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_grid(k.type.ss~bb.sp)

brm_representation3 <- brm(prop.visits ~ prop.flower.cover + (1 + prop.flower.cover | k.type.ss),
                 control = list(max_treedepth = 10),
                 #cores = 4,
                 cores = 2,
                 chains = 2,
                 #iter = 2000,
                 iter = 1000,
                 file = "./output/brm_representation3",
                 data = representation)

lmer_representation3 <- lmer(prop.visits ~ log(prop.flower.cover) + (1 + log(prop.flower.cover) | k.type.ss),
                             data = representation)

summary(lmer_representation3)
plot(lmer_representation3)
```

# 3. How do species vary in their biases?
```{r}
rep_matrix.plantsp <- representation %>%
  group_by(bb.sp, group) %>%
  summarize(bias = mean(bias)) %>%
  pivot_wider(values_from = bias, names_from = group, values_fill = 0) %>%
  column_to_rownames("bb.sp")

bias_matrix.ktype <- bias_group_ktype %>%
  group_by(bb.sp, group) %>%
  summarize(bias = mean(bias)) %>%
  pivot_wider(values_from = bias, names_from = group, values_fill = 0) %>%
  column_to_rownames("bb.sp")

traits <- bb_traits %>%
  arrange(bb.sp)

pca.plantsp <- rda(bias_matrix.plantsp)
plot(pca.plantsp, type = "t")
envfit(pca.plantsp ~ pbl.w.class, data = traits)
summary(ordisurf(pca.plantsp ~ pbl.w.class, data = traits))

pca.ktype <- rda(bias_matrix.ktype)
plot(pca.ktype, type = "t")
envfit(pca.ktype ~ pbl.w.class, data = traits)
summary(ordisurf(pca.ktype ~ pbl.w.class, data = traits))
pca.ktype_pca1_traits <- pca.ktype$CA$u[,1] %>%
  as.data.frame() %>%
  rownames_to_column(var = "bb.sp") %>%
  rename(PC1 = 2) %>%
  left_join(bb_traits)

ggplot(pca.ktype_pca1_traits, aes(pbl.w, PC1)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2))

ggplot(pca.ktype_pca1_traits, aes(pbl.w.class, PC1)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### 2021-01-13 writing/manuscript_v4.Rmd
restructured project directory; this will break some paths, but easy to fix, and clearer now

Plant-pollinator networks are characterized by unequal interaction strengths among partners. While the distribution of interaction strengths can be partially explained by the neutral effects of abundance, pollinators exhibit positive and negative biases toward available flora even when interaction strength is conditioned on floral abundance. This phenomenon is formalized by the discrimination (d’) metric of [@Bluthgen2006-if], but to avoid conflation with the metric *per se*, we will use the term bias in a more general sense to describe disproportionality between pollinator visitation and floral abundance.

Taken as a focal phenomenon, the realized floral biases of pollinators can be interposed between their underlying causal mechanisms and their higher-order significance **(Figure 1)**. From the causal perspective, realized biases can be partitioned into innate and learned components. Both these components depend on interactions between the functional traits of a given pollinator and those of its potential floral hosts, manifest over ecological time scales with respect to learned biases and evolutionary time scales with respect to innate biases. From the vantage of higher-order effects, the floral biases of pollinators contribute to the complex structures observed in plant-pollinator networks, which have consequences for both the ecological function of the network and the fitness of its constituent species. These effects can, in turn, feedback on the causes of bias via their modulation of community composition and biotic interactions.

Bumble bees (Hymenoptera: *Bombus* spp.) are an especially interesting group in which to study floral biases together with their causal mechanisms and downstream effects. On the one hand, bumble bees are exceptionally versatile foragers, and their ability to solve the puzzles posed by complex flower morphologies [@Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Even when a gross mismatch between tongue length and corolla depth makes the nectar of a given flower inaccessible by ordinary means, some bumble bee species will resort readily to biting through the base of the flower to “rob” nectar, incidentally making the nectar available to other prospective foragers as well. Flower handling skills are, moreover, socially transmitted, amplifying the plasticity of individual behavior [@Leadbeater2008-rh; @Loukola2017-bt]. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the searching/learning/handling costs associated with exploiting it. At an evolutionary level, the innate preferences of bumble bees for certain floral traits likely reflect historic associations between floral traits and floral rewards [@Raine2007-nq]. These innate preferences, however, can be modulated or even nullified by learned patterns of reward based on individual foraging experience [@Kugler1943-uy; Maharaj2019-wm], which depend not only intrinsic floral properties but also ecological modulators, such as the spatial distribution of floral patches or competition with other bumble bees [@Ishii2013-sm; @Plowright1984-lc; @Sowig1989-ld]. Thus, realized biases in bumble bee foraging are the labile product of bumble bee traits, floral traits, and ecological context. The influence of these biases on interaction network structure has consequences both for plant pollination and bumble bee species coexistence, and these consequences potentially feed back into the causes of bumble bee foraging bias via the modulation of competition between bumble bees and ultimately via the regulation of community assembly.  <!-- strengthen this sentence with some refs? -->

The topic of floral preference in bumble bees has been approached via both laboratory experimentation [e.g. @Kugler1943-uy; @Raine2007-nq] and observational field study [e.g. Heinrich1976-wc; @Goulson2008-zy]. The former approach enables more confident causal inference with respect to the mechanisms of floral choice, but it suffers all the usual limitations of biological realism associated with experimental control, and the higher-order effects of floral choices on network structure cannot be observed meaningfully under laboratory conditions. Observational field studies, however, have tended to be anecdotal in scope, typically describing bumble bee foraging patterns in particular localities, over short time frames, and with limited floristic data to enable a distinction between true foraging bias and the neutral effects of floral abundance [though see @Neumayer1999-il]. The steep elevation gradients of mountain slopes provide an opportunity for more powerful observational study of bumble foraging because they generate broad gradients of floristic and abiotic conditions while controlling for confounding effects of geographic variation, such as genotypic differentiation within bumble bee species. When bumble bee visitation in such environments is sampled with sufficient spatial and temporal replication and complemented with quantitative floral surveying, inference can be made regarding both the trait interactions that underly floral biases and the network structures that result from them. Mountain habitats are also arguably the most relevant places to study bumble bee ecology, since bumble bees are adapted to cool climates, and their global distribution is concentrated in areas of high elevation or high latitude. In such environments, bumble bees tend to dominate the pollinator community, and local flora often depend primarily on bumble bee visitation for pollination <!-- This paragraph needs to be beefed up a bit, but it at least sketches out the points I want to make -->.

@@

Prior to analysis, we aligned each bumble bee visitation sample with corresponding floral survey data. For most samples, floral surveying was conducted jointly with bumble bee visitation observation. In cases where bumble bee visitation was sampled without a joint floral survey, the bumble bee visitation data was paired with the nearest available floral survey date within a maximum distance of 7 days. Bumble bee visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases where more than one bumble bee visitation date was paired with a given floral survey date, the bumble bee visitation data were pooled.

@@

>>>>

Plant-pollinator networks are characterized by unequal interaction strengths among partners. While the distribution of interaction strengths can be partially explained by the neutral effects of abundance, pollinators exhibit positive and negative biases toward available flora even when interaction strength is conditioned on floral abundance. Taken as a focal phenomenon, the floral biases of pollinators can be interposed between their underlying causal mechanisms and their higher-order significance **(Figure 1)**. From the causal perspective, floral biases can be partitioned into innate and learned components. Both these components depend on interactions between the functional traits of a given pollinator and those of its potential floral hosts, manifest over ecological time scales with respect to learned biases and evolutionary time scales with respect to innate biases. From the vantage of higher-order effects, the floral biases of pollinators contribute to the complex structures observed in plant-pollinator networks, which have consequences for both the ecological function of the network and the fitness of its constituent species. These effects can, in turn, feedback on the causes of bias via their modulation of community composition and biotic interactions.

Bumble bees (Hymenoptera: *Bombus* spp.) are an especially interesting group in which to study floral biases together with their causal mechanisms and downstream effects. On the one hand, bumble bees are exceptionally versatile foragers, and their ability to solve the puzzles posed by complex flower morphologies [@Plowright1984-lc], or indeed by human investigators [@Alem2016-rn; @Loukola2017-bt], renders almost any flower potentially accessible. Even when a gross mismatch between tongue length and corolla depth makes the nectar of a given flower inaccessible by ordinary means, some bumble bee species will resort readily to biting through the base of the flower to “rob” nectar, incidentally making the nectar available to other prospective foragers as well. Flower handling skills are, moreover, socially transmitted, amplifying the plasticity of individual behavior [@Leadbeater2008-rh; @Loukola2017-bt]. Nevertheless, the flexibility of bumble bee foraging is tuned to floral reward, which depends on both the intrinsic reward of a given flower and the searching/learning/handling costs associated with exploiting it. At an evolutionary level, the innate preferences of bumble bees for certain floral traits likely reflect historic associations between floral traits and floral rewards [@Raine2007-nq]. These innate preferences, however, can be modulated or even nullified by patterns of reward learned during foraging [@Kugler1943-uy; Maharaj2019-wm], which depend not only intrinsic floral properties but also ecological modulators, such as the spatial distribution of floral patches or competition with other bumble bees [@Ishii2013-sm; @Plowright1984-lc; @Sowig1989-ld]. Thus, realized biases in bumble bee foraging are the labile product of bumble bee traits, floral traits, and ecological context. The influence of these biases on interaction network structure has consequences both for plant pollination and bumble bee species coexistence, and these consequences potentially feed back into the causes of bumble bee foraging bias via the modulation of competition between bumble bees and ultimately via the regulation of community assembly.  <!-- strengthen this sentence with some refs? -->

The topic of floral preference in bumble bees has been approached via both laboratory experimentation [e.g. @Kugler1943-uy; @Raine2007-nq] and observational field study [e.g. Heinrich1976-wc; @Goulson2008-zy]. The former approach enables more confident causal inference with respect to the mechanisms of floral choice, but it suffers all the usual limitations of biological realism associated with experimental control, and the higher-order effects of floral choices on network structure cannot be observed meaningfully under laboratory conditions. Observational field studies, however, have tended to be anecdotal in scope, typically describing bumble bee foraging patterns in particular localities, over short time frames, and with limited floristic data to enable a distinction between true foraging bias and the neutral effects of floral abundance [though see @Neumayer1999-il]. The steep elevation gradients of mountain slopes provide an opportunity for more powerful observational study of bumble foraging because they generate broad gradients of floristic and abiotic variation while controlling for confounding effects of geographic variation, such as genotypic differentiation within bumble bee species. When bumble bee visitation in such environments is sampled with sufficient spatial and temporal replication and complemented with quantitative floral surveying, inference can be made regarding both the trait interactions that underpin floral biases and the network structures that result from them. Mountain habitats are also arguably the most relevant places to study bumble bee ecology, since bumble bees are adapted to cool climates, and their global distribution is concentrated in areas of high elevation or high latitude. In such environments, bumble bees tend to dominate the pollinator community, and local flora often depend primarily on bumble bee visitation for pollination <!-- This paragraph needs to be beefed up a bit, but it at least sketches out the points I want to make -->.

@@

Prior to analysis, we aligned each bumble bee visitation sample with corresponding floral survey data. For most samples, floral surveying was conducted jointly with bumble bee visitation observation. In cases where bumble bee visitation was sampled without a joint floral survey, the bumble bee visitation data was paired with the nearest available floral survey data within a maximum distance of 7 days. Bumble bee visitation samples lacking corresponding floral survey data within 7 days were omitted from further analysis. In a few cases where more than one bumble bee visitation date was paired with a given floral survey date, the bumble bee visitation data were pooled.

@@

This phenomenon is formalized by the discrimination (d’) metric of @Bluthgen2006-if, but to avoid conflation with the metric *per se*, we use the term bias in a general sense to describe disproportionality between pollinator visitation and floral abundance.
